Dynamic Stochastic Approximation Monte Carlo

### 2018-07-23 papers/histogram/sad.tex
broad -> flat histogram and other edits.

\section{Introduction}
Over the past several decades a number of broad histogram Monte-Carlo
simulation algorithms have been developed which calculate the
thermodynamic properties of various systems for all temperatures.  The
development began with the original histogram method, which used a
single canonical Monte Carlo simulation to predict properties for
nearby temperatures~\cite{ferrenberg1988new}.  For large simulations
this approach is limited to a narrow temperature range because a single
canonical simulation explores only a small range of energies.  This led
to a variety of ``broad'' (or ``flat'') histogram
methods~\cite{penna1996broad, penna1998broad, swendsen1999transition,
wang2001determining, wang2001efficient, trebst2004optimizing}, which
attempt to explore a wide range of energies.  These methods also
benefit in that they are unlikely to be trapped in a low enetropy state.

@@

In this work, we have developed a novel algorithm based on SAMC that
does not require user-defined inputs and therefore should be easily
applicable to a given system.  We call this method SAD (Dynamic
Stochastic Approximation), and will discuss it in detail in the methods
section. We compare it along with several broad histogram methods
which include TMMC, WL, WL-TMMC, and SAMC.

@@

The organization of this paper is as follows: In Section~\ref{sec:histogram}, we
describe in detail broad-histogram methods.  Section~\ref{sec:weight} outlines
weight-based histogram methods used in this work. Section~\ref{sec:transition}
details the workings of the transition-matrix methods studied in this work.
Section~\ref{sec:sad} introduces the dynamic stochastic approximation method.
In Section~\ref{sec:results}, we discuss the performance of the various methods 
applied to the model system.

In this work, We compare a variety of broad histogram
methods.  We outline the general workings of each algorithm that we
developed in detail while summarizing algorithms that were developed in other
works.  The following methods are introduced and applied to the
square-well fluid: Transition Matrix Monte-Carlo (TMMC), Wang-Landau
(WL), Wang-Landau Transition Matrix Monte-Carlo (WL-TMMC), Stochastic
Approximation Monte-Carlo (SAMC), and Dynamic Stochastic Approximation
(SAD).

\section{Broad histogram methods}\label{sec:histogram}

The goal of broad histogram methods (also called \emph{flat histogram}
or \emph{multicanonical} methods) is to simulate each energy with
similar accuracy, so as to accurately determine the density of states
over a broad range of energies.  Once the density of states is known,
a number of thermodynamic quantities---such as heat capacity or
internal energy---can be easily computed for any temperature.
Properties that require more information---such as a spatial
correlation function or a response function---can still be computed
for any temperature, provided statistics are collected for each
individual energy, which can then be reweighted for any temperature.

Each broad histogram Monte Carlo method uses a given set of ``moves''
which change a system and should satisfy detailed balance.  Each
algorithm differs in how it determines the probability of accepting a
move, and in what additional statistics (if any) must be collected in
order decide on that probability.  The methods we will compare fall in
two categories:  weight-based methods, which only require a set of
weights $w(E)$ be collected, and transition-matrix methods that
require that statistics be collected for any transitions between pairs
of energies.

@@

We will begin by introducing two related weight-based methods which
rely on a weight function $w(E)$, noting that the method proposed in
this paper (discussed in Section~\ref{sec:sad}) also falls in this
category.  In these algorithms, the probability of accepting a move is
given by
\begin{equation}
	\mathcal{P}(\epsilon_\text{old} \rightarrow \epsilon_\text{new})
	= \min\left[1,\frac{w(\epsilon_\text{old})}{w(\epsilon_\text{new})}\right]
\end{equation}
the probability biases the simulation in favor of energies with low weights.  
A set of weights that are proportional to the density of states of the
system $D(\epsilon)$ will result in an entirely flat histogram.  Thus
flat histogram is a criteria for convergence for these methods.  To avoid 
overflow error , since the weights may vary over more than a
few hundred orders of magnitude, the natural logarithm of the weights are used 
in this work.  Since $D(\epsilon) = S(\epsilon)$, the logarithm of the weights 
can be thought of as an approximation of the entropy.

Each weighted density approach uses a random walk in energy space to
estimate the density of states.  The core of these approaches
is to continuously update the weights at each step of the simulation
\begin{equation}
	\ln{w_{t+1}(\epsilon)}=\ln{w_{t}(\epsilon)}
	+\gamma
\end{equation}
where $t$ is number of the current move, and $\gamma$ is an update
factor.  This update causes the random walk to avoid energies that
have been frequently sampled, leading to a rapid exploration of energy
space.  This approach, however, violates detailed balance, due to the
acceptance probabilities changing with each move.  The severity of
this violation decreased as we decrease $\gamma$.  The various
weight-based methods discussed here differ in how the decreasing of $\gamma$
is scheduled.

@@

\subsection{SAMC}
The Stochastic Approximation Monte Carlo (SAMC) algorithm addresses
the lack of convergence of Wang-Landau's approach with a simple
schedule by which the update factor $\gamma$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
  schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$ and in future works~\cite{werlich2015stochastic} with a scale factor
$\gamma_0$.
\begin{align}
\gamma_{t}^{\text{SA}} =\gamma_0 \frac{t_0}{\max(t_0,t)}
\end{align}
where as above $t$ is the number of moves that have been attempted.
The major advantage that SAMC offers is its proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has shown that the weights are proven
to converge to the true density of states~\cite{liang2006theory,
  liang2007stochastic}.  In addition, the energy range need not be
known a priori.  The time to converge depends only on the choice of
parameters $t_0$ and $\gamma_0$.  Unfortunately, there is no
prescription for finding acceptable values for these parameters, and
while the algorithm formally converges, for a poor choice of $t_0$
that convergence may be far too slow to be practical.  This work
proposes a means for dynamically setting the update factor $\gamma_t$
that should robustly converge on a variety of systems.

@@

\section{SAD Algorithm}\label{sec:sad}
\subsubsection{The theory behind SAD}
The SAD (Dynamic Stochastic Approximation) method is a version of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. Each
Broad-Histogram method has unique advantages and disadvantages.
Wang-Landau requires an energy range for initialization.  SAMC removes
this energy range requirement but requires simulating every possible
energy. Our proposed method SAD requires the user to input
$T_\text{min}$ which is an immediate disadvantage of the method;
however, this allows the user to define an energy simulation
range between $T_\text{min}$ and $\infty$. Setting a physical parameter
such as a minimum temperature $T_\text{min}$ is intuitive vs. a user
needing to know a prior either an energy range or some unphysical
parameter $t_0$.

While for SAMC the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SA}}$ can
be thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
$t_0$ should have dimensions of entropy.  We argue that we can write
this entropy `term' geometrically as an area bounded by the $\ln{D(E)}$
i.e. the entropy. This gives the following
\begin{align}
\gamma_{t}^{\text{SAD}} =
\frac{N_E}{t}\frac{\epsilon_{H}-\epsilon_{L}}{3T_{\text{min}}}
\end{align}
where $N_E$ is the number of energies that have been found. The change
in the entropy the minimum temperature can be thought of as a slope of
the triangle made by connecting the minimum important energy and the
maximum entropy state.  The factor of $3$ comes from the ratio of areas
bounded by the $\ln{D(E)}$ and the minimum temperature line. The update
factor has the same $1/t$ behavior as the original SAMC algorithm;
however now every time a new energy is discovered during the course of
the simulation, the update factor experiences a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly. In contrast to SAD, Fig.~\ref{fig:gamma-vs-t}. shows the
monotonic decrease of the update factor for SAMC.  The update factor
for WL decreases like a step function while the update factor for SAD
dynamically fluctuates around a value less than 1 for early MC moves
(while finding new energy states).  Ultimately once all energy states
have been found, the update factor for SAD decreases in the same
monotonic fashion as SAMC.

@@

Energy binning is an important consideration for complex continous
systems. Coarse binning yields fewer energy states for the walker to
explore but more time spent in the individual bin. Fine binning yields
more energy states to explore but less time spent in each bin.  Some
methods such as AdaWL~\cite{koh2013dynamically} (discussed elsewhere)
employ a tunable mechanism for controlling the binning for low entropic
states in order to ensure the exploration of all energies. A
significant advantage of SAD over other broad histogram methods is that
energy binning is solved dynamically. SAD should perform independently
for a reasonable range of binning because $\gamma \propto
\nicefrac{N_{E}}{t}$.  As the number of energy states found $N_{E}$
increases (fine binning), the time spent $t$ in each bin will decrease.
This argument explains why SAD should perform well for complex systems
in terms of binning. SAMC has a prefactor $\gamma_0$ to aid in a
similar way but this adds yet another parameter for the user to choose.

\subsubsection{Scaling, step size, and SAD}
Methods such as WL can fail to converge for complex systems in part
because the same random MC moves are performed for the entire energy
range.  The main problem lies in trying to determine how to optimize
the step size such that the time to explore all energies is reasonable.
Small step sizes lead to improved sampling at the expense of slow
convergence. Koh et al. proposed multiplying $\gamma$ by an energy
dependent factor in order to improve the converge of WL in regards to
convergence and scaling~\cite{koh2013dynamically}.  Because the update
factor for SAD has an energy dependent factor already present (and
dynamically adjusted), we expect that SAD will perform consitently
scaling free for a reasonable choice of step size.  SAD should perform
extremely well compared with other broad histogram methods when applied
to complex continuous systems such as polymers/proteins and spin models.

@@

J.K. Pommerenck would like to gratefully acknowledge D.W. Siderius and
J.R. Errington for helpful communications regarding the implementation
of WL-TMMC.

>>>>

\section{Introduction}
Over the past several decades a number of flat histogram Monte-Carlo
simulation algorithms have been developed which calculate the
thermodynamic properties of various systems for all temperatures.  The
development began with the original histogram method, which used a
single canonical Monte Carlo simulation to predict properties for
nearby temperatures~\cite{ferrenberg1988new}.  For large simulations
this approach is limited to a narrow temperature range because a single
canonical simulation explores only a small range of energies.  This led
to a variety of ``broad'' (or ``flat'') histogram
methods~\cite{penna1996broad, penna1998broad, swendsen1999transition,
wang2001determining, wang2001efficient, trebst2004optimizing}, which
attempt to explore a wide range of energies.  These methods also
benefit in that they are unlikely to be trapped in a low enetropy state.

@@

In this work, we have developed a novel algorithm based on SAMC that
does not require user-defined inputs and therefore should be easily
applicable to a given system.  We call this method SAD (Dynamic
Stochastic Approximation), and will discuss it in detail in the methods
section. We compare it along with several flat histogram methods
which include TMMC, WL, WL-TMMC, and SAMC.

@@

The organization of this paper is as follows: In Section~\ref{sec:histogram}, we
describe in detail flat-histogram methods.  Section~\ref{sec:weight} outlines
weight-based histogram methods used in this work. Section~\ref{sec:transition}
details the workings of the transition-matrix methods studied in this work.
Section~\ref{sec:sad} introduces the dynamic stochastic approximation method.
In Section~\ref{sec:results}, we discuss the performance of the various methods 
applied to the model system.

In this work, We compare a variety of flat histogram
methods.  We outline the general workings of each algorithm that we
developed in detail while summarizing algorithms that were developed in other
works.  The following methods are introduced and applied to the
square-well fluid: Transition Matrix Monte-Carlo (TMMC), Wang-Landau
(WL), Wang-Landau Transition Matrix Monte-Carlo (WL-TMMC), Stochastic
Approximation Monte-Carlo (SAMC), and Dynamic Stochastic Approximation
(SAD).

\section{Flat histogram methods}\label{sec:histogram}

The goal of flat histogram methods (also called \emph{broad histogram}
or \emph{multicanonical} methods) is to simulate each energy with
similar accuracy, so as to accurately determine the density of states
over a broad range of energies.  Once the density of states is known,
a number of thermodynamic quantities---such as heat capacity or
internal energy---can be easily computed for any temperature.
Properties that require more information---such as a spatial
correlation function or a response function---can still be computed
for any temperature, provided statistics are collected for each
individual energy, which can then be reweighted for any temperature.

Each flat histogram Monte Carlo method uses a given set of ``moves''
which change a system and should satisfy detailed balance.  Each
algorithm differs in how it determines the probability of accepting a
move, and in what additional statistics (if any) must be collected in
order decide on that probability.  The methods we will compare fall in
two categories:  weight-based methods, which only require a set of
weights $w(E)$ be collected, and transition-matrix methods that
require that statistics be collected for any transitions between pairs
of energies.

@@

We will begin by introducing two related weight-based methods which
rely on a weight function $w(E)$, noting that the method proposed in
this paper (discussed in Section~\ref{sec:sad}) also falls in this
category.  In these algorithms, the probability of accepting a move is
given by
\begin{equation}
	\mathcal{P}(E_\text{old} \rightarrow E_\text{new})
	= \min\left[1,\frac{w(E_\text{old})}{w(E_\text{new})}\right]
\end{equation}
the probability biases the simulation in favor of energies with low weights.  
A set of weights that are proportional to the density of states of the
system $D(E)$ will result in an entirely flat histogram.  Thus
flat histogram is a criteria for convergence for these methods.  To avoid 
overflow error , since the weights may vary over more than a
few hundred orders of magnitude, the natural logarithm of the weights are used
in this work.  Since $D(E) = S(E)$, the logarithm of the weights
can be thought of as an approximation of the entropy.

Each weighted density approach uses a random walk in energy space to
estimate the density of states.  The core of these approaches
is to continuously update the weights at each step of the simulation
\begin{equation}
	\ln{w_{t+1}(E)}=\ln{w_{t}(E)}
	+\gamma
\end{equation}
where $t$ is number of the current move, and $\gamma$ is an update
factor.  This update causes the random walk to avoid energies that
have been frequently sampled, leading to a rapid exploration of energy
space.  This approach, however, violates detailed balance, due to the
acceptance probabilities changing with each move.  The severity of
this violation decreased as we decrease $\gamma$.  The various
weight-based methods discussed here differ in how the decreasing of $\gamma$
is scheduled.

@@

\subsection{SAMC}
The Stochastic Approximation Monte Carlo (SAMC) algorithm addresses
the lack of convergence of Wang-Landau's approach with a simple
schedule by which the update factor $\gamma$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
  schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$.
\begin{align}
\gamma_{t}^{\text{SA}} =\frac{t_0}{\max(t_0,t)}
\end{align}
where as above $t$ is the number of moves that have been attempted.
The major advantage that SAMC offers is its proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has shown that the weights are proven
to converge to the true density of states~\cite{liang2006theory,
  liang2007stochastic}.  In addition, the energy range need not be
known a priori.  The time to converge depends only on the choice of
parameters $t_0$.  Unfortunately, there is no
prescription for finding an acceptable value for $t_0$, and
while the algorithm formally converges, for a poor choice of $t_0$
that convergence can be far too slow to be practical.

Werlich \emph{et al.} proposed scaling the SAMC $\gamma_t$ by a factor
$\gamma_0$.  While this may result in an improved rate of convergence,
it adds yet another parameter that must be empirically determined.

%This work proposes a means for dynamically setting the update factor
%$\gamma_t$ that should robustly converge on a variety of systems.

@@

\section{SAD Algorithm}\label{sec:sad}
\subsubsection{The theory behind SAD}
The SAD (Dynamic Stochastic Approximation) method is a version of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. Each
Flat-Histogram method has unique advantages and disadvantages.
Wang-Landau requires an energy range for initialization.  SAMC removes
this energy range requirement but requires simulating every possible
energy. Our proposed method SAD requires the user to input
$T_\text{min}$ which is an immediate disadvantage of the method;
however, this allows the user to define an energy simulation
range between $T_\text{min}$ and $\infty$. Setting a physical parameter
such as a minimum temperature $T_\text{min}$ is intuitive vs. a user
needing to know a prior either an energy range or some unphysical
parameter $t_0$.

While for SAMC the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SA}}$ can
be thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
$t_0$ should have dimensions of entropy.  We argue that we can write
this entropy `term' geometrically as an area bounded by the $\ln{D(E)}$
i.e. the entropy. This gives the following
\begin{align}
\gamma_{t}^{\text{SAD}} =
\frac{N_E}{t}\frac{E_{H}-E_{L}}{3T_{\text{min}}}
\end{align}
where $N_E$ is the number of energies that have been found. The change
in the entropy the minimum temperature can be thought of as a slope of
the triangle made by connecting the minimum important energy and the
maximum entropy state.  The factor of $3$ comes from the ratio of areas
bounded by the $\ln{D(E)}$ and the minimum temperature line. The update
factor has the same $1/t$ behavior as the original SAMC algorithm;
however now every time a new energy is discovered during the course of
the simulation, the update factor experiences a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly. In contrast to SAD, Fig.~\ref{fig:gamma-vs-t}. shows the
monotonic decrease of the update factor for SAMC.  The update factor
for WL decreases like a step function while the update factor for SAD
dynamically fluctuates around a value less than 1 for early MC moves
(while finding new energy states).  Ultimately once all energy states
have been found, the update factor for SAD decreases in the same
monotonic fashion as SAMC.

@@

Energy binning is an important consideration for complex continous
systems. Coarse binning yields fewer energy states for the walker to
explore but more time spent in each individual bin. Fine binning yields
more energy states to explore with less time spent in each bin.
\davidsays{move this discussion earlier.} Some
methods such as AdaWL~\cite{koh2013dynamically} (discussed elsewhere)
employ a tunable mechanism for controlling the binning for low entropic
states in order to ensure the exploration of all energies. A
significant advantage of SAD over other flat histogram methods is that
energy binning is solved dynamically. SAD should perform independently
for a reasonable range of binning because $\gamma \propto
\nicefrac{N_{E}}{t}$.  As the number of energy states found $N_{E}$
increases (fine binning), the time spent $t$ in each bin will decrease.
This argument explains why SAD should perform well for complex systems
in terms of binning. SAMC has a prefactor $\gamma_0$ to aid in a
similar way but this adds yet another parameter for the user to choose.

\subsubsection{Scaling, step size, and SAD}
Methods such as WL can fail to converge for complex systems in part
because the same random MC moves are performed for the entire energy
range.  The main problem lies in trying to determine how to optimize
the step size such that the time to explore all energies is reasonable.
Small step sizes lead to improved sampling at the expense of slow
convergence. Koh et al. proposed multiplying $\gamma$ by an energy
dependent factor in order to improve the converge of WL in regards to
convergence and scaling~\cite{koh2013dynamically}.  Because the update
factor for SAD has an energy dependent factor already present (and
dynamically adjusted), we expect that SAD will perform consitently
scaling free for a reasonable choice of step size.  SAD should perform
extremely well compared with other flat histogram methods when applied
to complex continuous systems such as polymers/proteins and spin models.

@@

We gratefully acknowledge D.W. Siderius and
J.R. Errington for helpful communications regarding the implementation
of WL-TMMC.
